---
title: "growth_curve_annotation"
author: "mc"
date: "2023-06-27"
output: html_document

knitr::opts_knit$set(root.dir = 'D:/Projects/cornell_ms_thesis/data/growth_curves/clean_data/annotated_data')
---

```{r setup, echo=TRUE, results='hide'}

library(dplyr)
library(reshape2)
library(ggplot2)
library(growthcurver)
library(purrr)
library(minpack.lm)
library(tidyr)

```

Logistic Model
N(t) = K / (1 + exp(-r *(t-t0)))

N(t)  - OD600 at time t
K     - carrying capacity (maximum OD600)
r     - growth rate
t0    - inflection point (time of maximum r)

Four Parameter Model
N(t) = A + ((B - A) / (1 + exp((t - C) / D)))

N(t)  - OD600 at time t
A     - lower asymptote, representing the initial value when t is very small
B     - upper asymptote, representing the final value when t is very large
C     - inflection point, which is the time of maximum growth rate
D     - related to the growth rate at the inflection point.

```{r functions, echo=TRUE, results='hide'}

#function that calculates 95% confidence intervals 
#for the given data vector using a t-distribution
conf_int95 <- function(data) {
  n <- length(data)
  error <- qt(0.975, df=n-1) * sd(data)/sqrt(n)
  return(error)
  }

#log-logistic model
logistic_model <- function(t, K, r, t0) {
  K / (1 + exp(-r*(t-t0)))
  }

#biphasic model, using double-logistic
biphasic_model <- function(t, K1, r1, t01, K2, r2, t02) {
  logistic_model(t, K1, r1, t01) + logistic_model(t, K2, r2, t02)
  }

#gompertz model
gompertz_model <- function(t, N0, umax, lambda) {
  N0 * exp(-exp((umax/exp(1)) * (lambda-t) + 1))
  }

```

```{r analysis, echo=TRUE, results='hide'}

annotated <- read.csv("all_annotated.csv") #read .csv data file

#combining read-data chunk and summary-stats into one
#%>% is the magrittr pipe operator, |> is the native Rstudio pipe
stats <- annotated |> 
  group_by(pH, Strain, Time) |> 
  summarise(N=length(OD600), 
            AverageOD=mean(OD600), 
            StDev=sd(OD600), 
            CI95=conf_int95(OD600)) |> 
  filter(!is.na(Strain))

```

```{r gg-plot, echo=TRUE}

graph_1 <- ggplot(data=stats, aes(x=Time/60, y=AverageOD, color=Strain)) + 
        geom_ribbon(aes(ymin=AverageOD-StDev, ymax=AverageOD+StDev, fill=Strain), color=NA, alpha=0.3) + #standard deviation
        #geom_ribbon(aes(ymin=AverageOD-CI95, ymax=AverageOD+CI95, fill=Strain), color=NA, alpha=0.3) + #95% confidence interval
        geom_line() + 
        #scale_y_log10() + 
        facet_grid(vars(Strain), vars(pH)) + 
        labs(x="Time (Hours)", y="Absorbance at 600 nm")

print(graph_1)

```

```{r all-models, echo=TRUE}
#log-logisitc fit
fit_log <- stats |> 
  group_by(Strain, pH) |> 
  do(fit = tryCatch(
    nlsLM(AverageOD ~ logistic_model(Time, K, r, t0), 
                 start = list(K = max(.$AverageOD, na.rm = TRUE), 
                              r = 0.1, 
                              t0 = max(.$Time, na.rm = TRUE)/2), 
          data = ., 
          control = nls.control(maxiter = 200)), 
    error = function(e) {
      message("Failed to fit model for ", .$Strain[1], " ", .$pH[1], ": ", e$message) 
      return(NULL)
      }
    ))

#biphasic fit
fit_biphasic <- stats |> 
  group_by(Strain, pH) |> 
  do(fit = tryCatch(
    nlsLM(AverageOD ~ biphasic_model(Time, K1, r1, t01, K2, r2, t02), 
          start = list(K1 = max(.$AverageOD, na.rm = TRUE)/2, 
                       r1 = 0.1, 
                       t01 = max(.$Time, na.rm = TRUE)/4, 
                       K2 = max(.$AverageOD, na.rm = TRUE)/2, 
                       r2 = 0.1, 
                       t02 = 3 * max(.$Time, na.rm = TRUE)/4), 
          data = ., 
          control = nls.control(maxiter = 200)),
    error = function(e) {
      message("Failed to fit model for ", .$Strain[1], " ", .$pH[1], ": ", e$message) 
      return(NULL)
      }
    ))

#gompertz fit
fit_gompertz <- stats |> 
  group_by(Strain, pH) |> 
  do(fit = tryCatch(
    nlsLM(AverageOD ~ gompertz_model(Time, N0, umax, lambda), 
          start = list(N0 = max(.$AverageOD, na.rm = TRUE), 
                       umax = 0.1, 
                       lambda = max(.$Time, na.rm = TRUE)/2), 
          data = ., 
          control = nls.control(maxiter = 200)),
    error = function(e) {
      message("Failed to fit model for ", .$Strain[1], " ", .$pH[1], ": ", e$message) 
      return(NULL)
      }
    ))

```

```{r all-predictions, echo=TRUE}

#generate time sequence predictions
#length.out = 240 indicates points every 12 minutes
#length.our = 144 indicates points every 20 minutes
time_s.pred <- seq(from = min(stats$Time, na.rm = TRUE), to = max(stats$Time, na.rm = TRUE), length.out = 576)

#generate predictions for the logistic model
predictions_logistic <- do.call(rbind, lapply(seq_len(nrow(fit_log)), function(i) {
  params <- coef(fit_log$fit[[i]]) 
  preds <- data.frame(
    Time = time_s.pred, 
    AverageOD = logistic_model(time_s.pred, params["K"], params["r"], params["t0"]), 
    Strain = fit_log$Strain[i], 
    pH = fit_log$pH[i], 
    Model = "Logistic"
    ) 
  return(preds)
  }))

#generate predictions for the biphasic model
predictions_biphasic <- do.call(rbind, lapply(seq_len(nrow(fit_biphasic)), function(i) {
  params <- coef(fit_biphasic$fit[[i]])
  preds <- data.frame(
    Time = time_s.pred, 
    AverageOD = biphasic_model(time_s.pred, params["K1"], params["r1"], params["t01"], params["K2"], params["r2"], params["t02"]), 
    Strain = fit_biphasic$Strain[i], 
    pH = fit_biphasic$pH[i], 
    Model = "Biphasic"
    ) 
  return(preds)
  }))

#generate predictions for the gompertz model
predictions_gompertz <- do.call(rbind, lapply(seq_len(nrow(fit_gompertz)), function(i) {
  params <- coef(fit_gompertz$fit[[i]]) 
  preds <- data.frame(
    Time = time_s.pred, 
    AverageOD = gompertz_model(time_s.pred, params["N0"], params["umax"], params["lambda"]), 
    Strain = fit_gompertz$Strain[i], 
    pH = fit_gompertz$pH[i], 
    Model = "Gompertz"
    ) 
  return(preds)
  }))

#combine the three sets of predictions
predictions <- rbind(predictions_logistic, predictions_biphasic, predictions_gompertz)

```

```{r all-graphs, echo=TRUE}

graph_a <- ggplot() + 
  geom_point(data = stats, aes(x = Time, y = AverageOD, color = )) + 
  geom_line(data = predictions, aes(x = Time, y = AverageOD, color = Model)) + 
  facet_grid(Strain ~ pH) + 
  theme_bw() + 
  labs(x = "Time", y = "OD600", color = "Model")

print(graph_a)

```

```{r stats-functions, echo=TRUE}

#compute lag-time
lag_time <- function(fit, model_type) {
  coef <- coef(fit) 
  switch(model_type,
         "logistic" = {log(2) / coef['r']},
         "biphasic" = {min(log(2)/coef['r1'], log(2)/coef['r2'])},
         "gompertz" = {coef['lambda']}
  )/60
  }

#compute growth rate
growth_rate <- function(fit, model_type) {
  coef <- coef(fit) 
  switch(model_type, 
         "logistic" = {coef['K']}, 
         "biphasic" = {max(coef['K1'], coef['K2'])}, 
         "gompertz" = {coef['umax']}
  )
  }


#compute initial OD value (y0)
initial_od <- function(fit, model_type) {
  coef <- coef(fit) 
  switch(model_type, 
         "logistic" = {coef['K'] / (1 + exp(coef['r']*coef['t0']))}, 
         "biphasic" = {coef['K1'] / (1 + exp(coef['r1']*coef['t01'])) + coef['K2'] / (1 + exp(coef['r2']*coef['t02']))}, 
         "gompertz" = {coef['N0'] * exp(-exp((coef['umax']/exp(1)) * (coef['lambda']) + 1))}
  )
  }

#compute C value (increase in absorbance from y0 to ymax)
compute_c <- function(fit, model_type) {
  coef <- coef(fit) 
  switch(model_type, 
         "logistic" = {coef['K'] - initial_od(fit, model_type)}, 
         "biphasic" = {max(coef['K1'], coef['K2']) - initial_od(fit, model_type)}, 
         "gompertz" = {coef['N0'] * exp(coef['umax'] * (coef['lambda'] - 1/exp(1))) - initial_od(fit, model_type)}
  )
  }

#compute ymax
ymax <- function(fit, model_type) {
  coef <- coef(fit) 
  switch(model_type, 
         "logistic" = {coef['K']}, 
         "biphasic" = {max(coef['K1'], coef['K2'])}, 
         "gompertz" = {coef['N0'] * exp(-exp((coef['umax']/exp(1)) * (coef['lambda']) + 1))}
  )
  }

#compute sse
compute_sse <- function(fit) {
  residuals <- resid(fit)
  sse <- sum(residuals^2)
  return(sse)
  }

#compute r^2
compute_r2 <- function(fit, data) {
  residuals <- resid(fit) 
  sse <- sum(residuals^2) 
  sst <- sum((data$AverageOD - mean(data$AverageOD))^2) 
  r_squared <- 1 - sse/sst 
  return(r_squared)
  }

#compute rmse
compute_rmse <- function(fit) {
  residuals <- resid(fit) 
  rmse <- sqrt(mean(residuals^2))
  return(rmse)
  }

```

```{r stats-apply, echo=TRUE}

#compute statistics for each model

#logistic
summary_log <- fit_log |>
  rowwise() |> 
  mutate(lag_time = lag_time(fit, "logistic"), 
         growth_rate = growth_rate(fit, "logistic"), 
         initial_od = initial_od(fit, "logistic"), 
         c_value = compute_c(fit, "logistic"), 
         ymax = ymax(fit, "logistic"), 
         SSE = compute_sse(fit), 
         Rsquared = compute_r2(fit, data = stats), 
         RMSE = compute_rmse(fit))

#biphasic
summary_biphasic <- fit_biphasic |> 
  rowwise() |> 
  mutate(lag_time = lag_time(fit, "biphasic"), 
         growth_rate = growth_rate(fit, "biphasic"), 
         initial_od = initial_od(fit, "biphasic"), 
         c_value = compute_c(fit, "biphasic"), 
         ymax = ymax(fit, "biphasic"), 
         SSE = compute_sse(fit), 
         Rsquared = compute_r2(fit, data = stats), 
         RMSE = compute_rmse(fit))

#gompertz
summary_gompertz <- fit_gompertz |> 
  rowwise() |> 
  mutate(lag_time = lag_time(fit, "gompertz"), 
         growth_rate = growth_rate(fit, "gompertz"), 
         initial_od = initial_od(fit, "gompertz"), 
         c_value = compute_c(fit, "gompertz"), 
         ymax = ymax(fit, "gompertz"), 
         SSE = compute_sse(fit), 
         Rsquared = compute_r2(fit, data = stats), 
         RMSE = compute_rmse(fit))

summary_log$model <- "logistic"
summary_biphasic$model <- "biphasic"
summary_gompertz$model <- "gompertz"

# combine all model statistics
summary_all <- rbind(summary_log, summary_biphasic, summary_gompertz)

```

```{r model-decide, echo=TRUE}

best_model <- summary_all |>
  group_by(pH, Strain) |>
  summarise(
    best_RMSE_model = model[which.min(RMSE)],
    min_RMSE = min(RMSE),
    best_Rsquared_model = model[which.max(Rsquared)],
    max_Rsquared = max(Rsquared)
    )

```



```{r stats-apply, echo=TRUE}



```












From here we select models we want to use for specific quadrants
```{r log-modeling, echo=TRUE}

subset_log <- stats |> 
  filter(
      (pH == 4.6) | 
      (Strain == "9" & pH %in% c(5.8, 6.1)) | 
      (Strain == "B-768" & pH != 7.2) | 
      (Strain == "NRS-609" & pH %in% c(4.9, 5.5))
      )

fit_log <- subset_log |> 
  group_by(Strain, pH) |> 
  do(fit = nlsLM(AverageOD ~ logistic_model(Time, K, r, t0), 
                 start = list(K = max(.$AverageOD, na.rm = TRUE), r = 0.1, t0 = max(.$Time, na.rm = TRUE)/2), 
                 data = .)
     )

```

```{r biphasic-modeling, echo=TRUE}

subset_biphasic <- stats |> 
  filter(
      (Strain == "9" & pH %in% c(7.2, 5.2, 4.9)) | 
      (Strain == "11" & !(pH %in% c(4.6, 6.4, 7.2))) | 
      (Strain == "15" & pH %in% c(4.9, 5.2, 5.5)) | 
      (Strain == "16" & pH %in% c(5.8, 5.5, 5.2, 4.9)) | 
      (Strain == "NRS-609" & !(pH %in% c(7.2, 5.5, 4.6, 4.9)))
      )


fit_biphasic <- subset_biphasic |> 
  group_by(Strain, pH) |> 
  do(fit = nlsLM(AverageOD ~ biphasic_model(Time, K1, r1, t01, K2, r2, t02), 
                 start = list(K1 = max(.$AverageOD, na.rm = TRUE)/2, r1 = 0.1, t01 = max(.$Time, na.rm = TRUE)/4,
                              K2 = max(.$AverageOD, na.rm = TRUE)/2, r2 = 0.1, t02 = 3*max(.$Time, na.rm = TRUE)/4), 
                 data = ., control = nls.control(maxiter = 100))
     )

```

```{r predictions-model, echo=TRUE}

#generate time sequence predictions
#length.out = 240 indicates points every 12 minutes
#length.our = 144 indicates points every 20 minutes
time_s.pred <- seq(from = min(stats$Time, na.rm = TRUE), to = max(stats$Time, na.rm = TRUE), length.out = 576)

#generate predictions for the logistic model
predictions_logistic <- do.call(rbind, lapply(seq_len(nrow(fit_log)), function(i) {
  params <- coef(fit_log$fit[[i]]) 
  preds <- data.frame(
    Time = time_s.pred, 
    AverageOD = logistic_model(time_s.pred, params["K"], params["r"], params["t0"]), 
    Strain = fit_log$Strain[i], 
    pH = fit_log$pH[i], 
    Model = "Logistic"
    ) 
  return(preds)
  }))

#generate predictions for the biphasic model
predictions_biphasic <- do.call(rbind, lapply(seq_len(nrow(fit_biphasic)), function(i) {
  params <- coef(fit_biphasic$fit[[i]])
  preds <- data.frame(
    Time = time_s.pred, 
    AverageOD = biphasic_model(time_s.pred, params["K1"], params["r1"], params["t01"], params["K2"], params["r2"], params["t02"]), 
    Strain = fit_biphasic$Strain[i], 
    pH = fit_biphasic$pH[i], 
    Model = "Biphasic"
    ) 
  return(preds)
  }))

#combine the two sets of predictions
predictions <- rbind(predictions_logistic, predictions_biphasic)

```

```{r graphing-model, echo=TRUE}

graph_l <- ggplot() + 
  geom_point(data = stats, aes(x = Time, y = AverageOD, color = )) + 
  geom_line(data = predictions, aes(x = Time, y = AverageOD, color = Model)) + 
  facet_grid(Strain ~ pH) + 
  theme_bw() + 
  labs(x = "Time", y = "OD600", color = "Model")

print(graph_l)

```










